---
title: "Introduction to xafty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to xafty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
The goal of xafty is to build R-based pipelines that are reusable, scalable, and easy to collaborate on. To achieve this end, xafty represents your pipeline as a directed acyclic graph (DAG) where each step becomes a node in a network. 

This changes how you work with pipelines in one significant way: instead of executing steps in a predetermined sequence, xafty dynamically constructs the pipeline at run time based on the data you request.

This article gives you an intro to building network-pipelines. If you’ve already put together pipelines in R using base R or dplyr, you’ve got everything you need to start building them with xafty.

```{r setup}
library(xafty)
```

To create a network object, we call the `init_network()` function.

``` r
network <- init_network("intro_network")
```
Each network object must have at least one project in order to start building the network pipeline. 

A project can be added as following:

``` r
network$add_project("mtcars")
```
Projects are multi-purpose structures which are used to hold and organize your nodes, as well as encapsulate logic and settings how nodes should be treated within their respective projects.

This is all the setup we need, to start building our network.
